---
title: "cores"
author: "Jovan Tanevski"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html:
    df_print: kable
editor_options:
  chunk_output_type: console
---

## Setup

```{r setup, message = FALSE}
library(tidyverse)
library(skimr)
library(uwot)
library(limma)
library(NMF)
library(factoextra)
library(cowplot)
library(pheatmap)
library(RColorBrewer)
```

```{r}
data <- read_csv("data/tumor_hepatocytes.csv", col_types = cols())
tumor.hc <- data %>%
  select(
    `Cytoplasm AGS (Opal 690) Mean (Normalized Counts, Total Weighting)`,
    `Cytoplasm BerEP4 (Opal 650) Mean (Normalized Counts, Total Weighting)`,
    `Cytoplasm CRP (Opal 540) Mean (Normalized Counts, Total Weighting)`,
    `Nucleus p-S6 (Opal 570) Mean (Normalized Counts, Total Weighting)`,
#    `Nucleus beta-cat. (Opal 520) Mean (Normalized Counts, Total Weighting)`
  ) %>%
  `colnames<-`(str_split(colnames(.), " ") %>% map_chr(~ .x[2]) %>% make.names())

skim(tumor.hc)
```

Detect outliers based on Tukey's interquartile approach and winsorize. Follow by quantile normalization and ranking to get rid of the effect of abundance

```{r}
quartiles <- apply(tumor.hc, 2, \(x) quantile(x, c(.25, .75)))
lower <- quartiles[1, ] - 1.5 * (quartiles[2, ] - quartiles[1, ])
upper <- quartiles[2, ] + 1.5 * (quartiles[2, ] - quartiles[1, ])


tumor.hc.winsorized <- tumor.hc %>% imap_dfc(\(x, y){
  x[x < lower[y]] <- x[which.min(abs(x - lower[y]))]
  x[x > upper[y]] <- x[which.min(abs(x - upper[y]))]
  x
})

skim(tumor.hc.winsorized)

tumor.hc.norm <- normalizeQuantiles(data.frame(tumor.hc.winsorized))

skim(tumor.hc.norm)

tumor.hc.rank <- mutate_all(tumor.hc.winsorized, ~ rank(., ties.method = "min"))
```

Pilot run. Subsample 10% of the original data.

```{r}
set.seed(42)
percent <- 10
subsamp <- sample(nrow(tumor.hc.rank), (percent / 100) * nrow(tumor.hc.rank))
```


## Dimensionality reduction

```{r}
cache <- "output/tumor.hc.umap.rds"
if (file.exists(cache)) {
  tumor.hc.umap <- read_rds(cache)
} else {
  tumor.hc.umap <- umap(tumor.hc.norm, n_neighbors = 100, 
                        min_dist = 0.2, n_threads = 7)
  write_rds(tumor.hc.umap, cache, "gz")
}
```

Check if sample is representative in UMAP space

```{r}
tumor.hc.umap.sample <-
  tumor.hc.umap %>%
  `colnames<-`(c("U1", "U2")) %>%
  as_tibble()

all <- ggplot(tumor.hc.umap.sample, aes(x = U1, y = U2)) +
  geom_point(size = 0.5) +
  theme_classic()

sampled <- ggplot(tumor.hc.umap.sample %>% slice(subsamp), aes(x = U1, y = U2)) +
  geom_point(color = "darkgreen", size = 0.5) +
  theme_classic()

unsampled <- ggplot(tumor.hc.umap.sample %>% slice(-subsamp), aes(x = U1, y = U2)) +
  geom_point(color = "darkred", size = 0.5) +
  theme_classic()

plot_grid(all, sampled, unsampled)
```


## Consensus NMF

We use an efficient implementation of alternating non negative least-squares with regularized to favor sparse coefficient matrices snmf/r. In this way we aim for cleaner clustering.

```{r}
cache <- paste0("output/tumor.hc.nmf.rank.", percent, ".rds")

if (file.exists(cache)) {
  tumor.hc.nmf.rank <- read_rds(cache)
} else {
  tumor.hc.nmf.rank <- nmfEstimateRank(as.matrix(t(tumor.hc.rank[subsamp, ])),
    range = seq(2, 5), method = "snmf/r",
    nrun = 10, seed = 42, verbose = TRUE,
    .options = "mp5"
  )
  write_rds(tumor.hc.nmf.rank, cache, "gz")
}

plot(tumor.hc.nmf.rank)
```
