---
title: "cores"
author: "Jovan Tanevski"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html:
    df_print: kable
editor_options:
  chunk_output_type: console
---

## Setup

```{r setup, message = FALSE}
library(skimr)
library(uwot)
library(limma)
library(NMF)
library(cowplot)
library(pheatmap)
library(RColorBrewer)
library(distances)
library(furrr)
library(raster)
library(RStoolbox)
library(tidyverse)
```

```{r}
data <- read_csv("data/tumor_hepatocytes.csv", col_types = cols())
tumor.hc <- data %>%
  select(
    `Cytoplasm AGS (Opal 690) Mean (Normalized Counts, Total Weighting)`,
    `Cytoplasm BerEP4 (Opal 650) Mean (Normalized Counts, Total Weighting)`,
    `Cytoplasm CRP (Opal 540) Mean (Normalized Counts, Total Weighting)`,
    `Nucleus p-S6 (Opal 570) Mean (Normalized Counts, Total Weighting)`,
#    `Nucleus beta-cat. (Opal 520) Mean (Normalized Counts, Total Weighting)`
  ) %>%
  `colnames<-`(str_split(colnames(.), " ") %>% map_chr(~ .x[2]) %>% make.names())

skim(tumor.hc)
```

Detect outliers based on Tukey's interquartile approach and winsorize. Follow by quantile normalization and ranking to get rid of the effect of abundance

```{r}
quartiles <- apply(tumor.hc, 2, \(x) quantile(x, c(.25, .75)))
lower <- quartiles[1, ] - 1.5 * (quartiles[2, ] - quartiles[1, ])
upper <- quartiles[2, ] + 1.5 * (quartiles[2, ] - quartiles[1, ])


tumor.hc.winsorized <- tumor.hc %>% imap_dfc(\(x, y){
  x[x < lower[y]] <- x[which.min(abs(x - lower[y]))]
  x[x > upper[y]] <- x[which.min(abs(x - upper[y]))]
  x
})

skim(tumor.hc.winsorized)

tumor.hc.norm <- normalizeQuantiles(data.frame(tumor.hc.winsorized))

skim(tumor.hc.norm)

tumor.hc.rank <- mutate_all(tumor.hc.winsorized, ~ rank(., ties.method = "min"))
```

Pilot run. Subsample one core per patient from the original data.

```{r}
set.seed(42)
selected.cores <- data %>% select(`Sample Name`) %>% 
  mutate(sample = str_extract(`Sample Name`, "[0-9]{2}_[0-9]+")) %>% 
  group_by(`Sample Name`) %>% distinct() %>% ungroup() %>% 
  group_by(sample) %>% slice_sample() %>% pull(`Sample Name`)

subsamp <- which(data %>% pull(`Sample Name`) %in% selected.cores)
```


## Dimensionality reduction

```{r}
cache <- "output/tumor.hc.umap.rds"
if (file.exists(cache)) {
  tumor.hc.umap <- read_rds(cache)
} else {
  tumor.hc.umap <- umap(tumor.hc.norm, n_neighbors = 100, 
                        min_dist = 0.2, n_threads = 7)
  write_rds(tumor.hc.umap, cache, "gz")
}

tumor.hc.umap.sample <-
  tumor.hc.umap %>%
  `colnames<-`(c("U1", "U2")) %>%
  as_tibble()

```

Check if sample is representative in UMAP space

```{r}
all <- ggplot(tumor.hc.umap.sample, aes(x = U1, y = U2)) +
  geom_point(size = 0.5) +
  theme_classic()

sampled <- ggplot(tumor.hc.umap.sample %>% slice(subsamp), aes(x = U1, y = U2)) +
  geom_point(color = "darkgreen", size = 0.5) +
  theme_classic()

unsampled <- ggplot(tumor.hc.umap.sample %>% slice(-subsamp), aes(x = U1, y = U2)) +
  geom_point(color = "darkred", size = 0.5) +
  theme_classic()

plot_grid(all, sampled, unsampled)
```


## Consensus NMF

We use an efficient implementation of alternating non negative least-squares with regularized to favor sparse coefficient matrices snmf/r. In this way we aim for cleaner clustering.

```{r}
cache <- "output/tumor.hc.nmf.all.3.rds"

if (file.exists(cache)) {
  tumor.hc.nmf <- read_rds(cache)
} else {
  tumor.hc.nmf <- nmf(as.matrix(t(tumor.hc.rank[subsamp, ])),
    rank = 3, method = "snmf/r",
    nrun = 10, seed = 42, verbose = TRUE,
    .options = "vkp10-m"
  )
  write_rds(tumor.hc.nmf, cache, "gz")
}
```

Extract basis of NMF (signature of cluster)

```{r}
basismap(tumor.hc.nmf)
```

Assign clusters
```{r}
nmf.clusters <- apply(fit(tumor.hc.nmf)@H, 2, which.max)
```

### Plot in 2D

```{r}
tumor.hc.umap.clus <-
  tumor.hc.umap.sample %>%
  slice(subsamp) %>%
  mutate(Cluster = as.factor(nmf.clusters))

ggplot(tumor.hc.umap.clus, aes(x = U1, y = U2, color = Cluster)) +
  geom_point(size = 0.5) +
  theme_classic()
```

### Expression profiles per cluster 

```{r fig.width=10, fig.height=10}
tumor.hc.clustered.nmf <- tumor.hc.norm[subsamp, ] %>%
  mutate(Cluster = as.factor(nmf.clusters)) %>%
  pivot_longer(names_to = "Marker", values_to = "Norm.value", -Cluster)

profiles <- seq_len(max(nmf.clusters)) %>% map(~
ggplot(
  tumor.hc.clustered.nmf %>% filter(Cluster == .x),
  aes(x = Marker, y = Norm.value, color = Marker)
) +
  stat_summary(fun.data = mean_sdl, show.legend = FALSE) +
  scale_color_brewer(palette = "Set2") +
  ylim(-1, 3) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)))

plot_grid(plotlist = profiles, labels = paste("Cluster", seq_len(max(nmf.clusters))))
```

### Marker abundance plots

```{r}
tumor.hc.umap.markers <- tumor.hc.norm %>%
  bind_cols(tumor.hc.umap.sample) %>%
  slice(subsamp)

low <- RColorBrewer::brewer.pal(8, "Set2")[8]
highs <- RColorBrewer::brewer.pal(8, "Set2")[seq_len(ncol(tumor.hc.norm))]

tumor.hc.umap.markers.plots <- colnames(tumor.hc.norm) %>%
  map2(highs, \(marker, color){
    ggplot(tumor.hc.umap.markers, aes_string(x = "U1", y = "U2", color = marker)) +
      geom_point(size = 0.5) +
      scale_color_gradient(low = low, high = color) +
      theme_classic()
  })

plot_grid(plotlist = tumor.hc.umap.markers.plots)
```

### Core plots

```{r}
tumor.hc.umap.cores <- data %>%
  select(`Sample Name`) %>%
  bind_cols(tumor.hc.umap.sample) %>%
  slice(subsamp) %>%
  mutate(
    c = nmf.clusters,
    sample = str_extract(`Sample Name`, "[0-9]+_[0-9]+")
  )

tumor.hc.umap.cores %>%
  group_by(sample) %>%
  summarize(
    Fraction = table(c) / n(),
    Cluster = names(Fraction),
    .groups = "drop"
  ) %>%
  mutate(Fraction = as.numeric(Fraction)) %>%
  pivot_wider(names_from = "Cluster", values_from = "Fraction") %>%
  column_to_rownames("sample") %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>%
  as.matrix() %>%
  pheatmap(
    scale = "none",
    color = colorRampPalette(brewer.pal(n = 7, name = "YlOrBr"))(100),
    fontsize = 6
  )

tumor.hc.umap.cores %>%
  pull(sample) %>%
  unique() %>%
  walk(\(s){
    output.fig <- paste0("output/cores_3/", s, ".png")
    if (!file.exists(output.fig)) {
      png(output.fig, width  = 800, height = 800)
      (ggplot(
        tumor.hc.umap.cores %>%
          mutate(c = ifelse(sample == s, c, NA), Cluster = as.factor(c)) %>%
          arrange(!is.na(Cluster), Cluster),
        aes(x = U1, y = U2, color = Cluster)
      ) +
        geom_point(size = 0.5) +
        scale_color_discrete(na.value = "gray80") +
        theme_classic()) %>%
        print()
      dev.off()
    }
  })

```

Figures with UMAPs for each core can be found in [output](https://github.com/saezlab/MultispectralHCC/tree/master/output).


Overlay cluster information on available tiffs
```{r}
available.images <- list.files("data/core images/", full.names = TRUE)

spatial <- data %>% 
  select(`Sample Name`, `Cell X Position`, `Cell Y Position`) %>% 
  `colnames<-`(c("sample", "X", "Y")) %>%
  slice(subsamp) %>% 
  mutate(Cluster = as.factor(nmf.clusters))

available.images %>% walk(\(img){
  id <- str_extract(img, "[0-9]{2}_[0-9]+(_[^_\\.]*){4}")
  name <- str_extract(img, "[0-9]{2}_[0-9]+(_[^_\\.]*)*.tif")
  s <- paste0(id,".im3")
  if(s %in% spatial$sample){
    rb <- brick(img)
    
    
    names(rb) <- c("r", "g", "b")
    
    pdf(paste0("output/cores_3/", name, ".pdf"))
    #the t should be flipped along the y direction to match coordinates in spatial
    (ggRGB(flip(rb, "y")) + 
      geom_point(data = spatial %>% filter(sample == s), aes(x = X, y = Y, color = Cluster)) +
      theme_map()) %>%
      print()
    dev.off()
  }
  
})

```


### Differential expression analysis (silhouette)

Calculate the similarity of samples using the expression and the silhouette scores based on the assigned clusters.
```{r}
cache <- "output/silhouette.nmf.rds"

if (file.exists(cache)) {
  silhouette.nmf <- read_rds(cache)
} else {
  # manual calculation of silhouette scores with lazily evaluated distance matrix
  subsamp.dists <- distances(tumor.hc.norm[subsamp, ])
  
  plan(multisession, workers = 5)
  
  silhouette.nmf <- nmf.clusters %>% future_imap_dbl(\(c, i){
    dists <- tibble(d = subsamp.dists[i,][-i], cluster = nmf.clusters[-i]) %>% 
      group_by(cluster) %>% 
      summarize(m = mean(d)) 
    
    a <- dists %>% filter(cluster == c) %>% pluck("m", 1)
    b <- dists %>% filter(cluster != c) %>% pull(m) %>% min()
    
    (b - a)/max(a, b)
  }, .options = furrr::furrr_options(packages = "distances"), .progress = TRUE)
  
  write_rds(silhouette.nmf, cache, "gz")
}

tibble(c = nmf.clusters, s = silhouette.nmf) %>% 
  group_by(c) %>% 
  summarize(m = mean(s))

print(paste0("Average silhouette score: ", mean(silhouette.nmf)))
```

Select only the samples with positive silhouette scores as "core samples"
```{r}
core.samples <- which(silhouette.nmf > 0)
tumor.hc.core.samples <- tumor.hc.norm[subsamp, ] %>%
  add_column(Cluster = nmf.clusters) %>%
  slice(core.samples)
```

Calculate difference in means (mean(cluster) - mean(other)), one-vs-all t-test per marker and correct for FDR. Filter q <= 0.05. Plot the differences.

```{r}
de.table <- unique(tumor.hc.core.samples$Cluster) %>%
  map_dfr(\(c){
    tumor.hc.core.samples %>%
      summarize(across(-Cluster, ~ t.test(.x ~ (Cluster == c))$p.value)) %>%
      pivot_longer(names_to = "Marker", values_to = "p", everything()) %>%
      mutate(Cluster = c, Difference = tumor.hc.core.samples %>%
        group_by(Cluster == c) %>%
        select(-Cluster) %>%
        group_split(.keep = FALSE) %>% map(colMeans) %>% reduce(`-`))
  }) %>%
  mutate(q = p.adjust(p, method = "fdr"), Difference = -Difference)

de.table %>%
  filter(q <= 0.05) %>%
  arrange(q)

de.table %>%
  pivot_wider(names_from = "Cluster", values_from = "Difference", -c(p, q)) %>%
  column_to_rownames("Marker") %>%
  as.matrix() %>%
  pheatmap(scale = "none")
```
