---
title: "Analysis of TCGA RPPA LIHC samples"
author: "Jovan Tanevski"
date: "`r Sys.Date()`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r include = FALSE}
knitr::opts_chunk$set(dev = "svg")
```


## Setup

Load required libraries.

```{r setup, message = FALSE}
library(tidyverse)
library(skimr)
library(uwot)
library(factoextra)
library(cowplot)
library(FNN)
library(igraph)
library(leiden)
library(limma)
library(NMF)
```

Read filtered TCGA RRPA data and display summary statistics.

```{r}
tcga.raw <- read_csv("data/TCGA-RPPA-LIHC_selected.csv", col_types = cols()) %>%
  select(-TumorType) %>%
  column_to_rownames("SampleID")

skim(tcga.raw)
```

Rank normalize as suggested in https://gdac.broadinstitute.org/runs/analyses__2016_01_28/reports/cancer/LIHC/RPPA_Clustering_CNMF/nozzle.html

```{r}
tcga <- mutate_all(tcga.raw, rank)
```


## Dimensionality reduction

```{r}
tcga.pca <- prcomp(tcga)
summary(tcga.pca)

tcga.umap <- umap(tcga, n_neighbors = 5, n_epochs = 1000)
```


## Hierarchical clustering

Perform hierarchical clustering of the data and plot the resulting dendrogram

```{r fig.width=16, fig.height=9}
tcga.hclust <- eclust(tcga, "hclust")

fviz_dend(tcga.hclust, rect = TRUE)

fviz_gap_stat(tcga.hclust$gap_stat)

fviz_silhouette(tcga.hclust) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Plot in 2D PCA and UMAP

```{r}
fviz_pca_ind(tcga.pca, geom = "point", col.ind = as.factor(tcga.hclust$cluster)) +
  theme_classic()

tcga.umap.clus <-
  tcga.umap %>%
  cbind(tcga.hclust$cluster) %>%
  `colnames<-`(c("U1", "U2", "Cluster")) %>%
  as_tibble() %>%
  mutate_at("Cluster", as.factor)

ggplot(tcga.umap.clus, aes(x = U1, y = U2, color = Cluster)) +
  geom_point() +
  theme_classic()
```

Expression profiles per cluster 

```{r fig.width=10, fig.height=10}
tcga.clustered <- tcga.raw %>%
  mutate(Cluster = as.factor(tcga.hclust$cluster)) %>%
  pivot_longer(names_to = "Marker", values_to = "Z", -Cluster)

profiles <- seq_len(tcga.hclust$nbclust) %>% map(~
ggplot(
  tcga.clustered %>% filter(Cluster == .x),
  aes(x = Marker, y = Z, color = Marker)
) +
  stat_summary(fun.data = mean_sdl, show.legend = FALSE) +
  ylim(-3, 3) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)))

plot_grid(plotlist = profiles, labels = paste("Cluster", seq_len(tcga.hclust$nbclust)))
```


Differential expression analysis

```{r}
design <- model.matrix(~0 + as.factor(tcga.hclust$cluster))
colnames(design) <- paste0("Cluster", seq_len(tcga.hclust$nbclust))

tcga.hclust.limma <- lmFit(t(tcga.raw), design = design)

tcga.hclust.limma$coefficients

tests.hclust <- decideTests(eBayes(tcga.hclust.limma))

tests.hclust@.Data

summary(tests.hclust)

```


## Similarity and graph based clustering

Calculate leiden clustering based on weighted shared nearest neighbor graph
```{r}
tcga.knn <- knn.index(tcga, 10)
jaccard.cuttoff <- 0.1

snn <- seq_len(nrow(tcga.knn) - 1) %>% map_dfr(\(id){
  to <- seq(id + 1, nrow(tcga.knn))
  jaccard <- to %>%
    map_dbl(~ length(intersect(tcga.knn[id, ], tcga.knn[.x, ])) /
      length(union(tcga.knn[id, ], tcga.knn[.x, ])))
  
  tibble(from = id, to = to, weight = jaccard) %>% filter(jaccard >= jaccard.cuttoff)
})

leiden.clusters <- graph_from_data_frame(snn, directed = FALSE) %>% leiden()

```

Plot in 2D PCA and UMAP

```{r}
fviz_pca_ind(tcga.pca, geom = "point", col.ind = as.factor(leiden.clusters)) +
  theme_classic()

ggplot(tcga.umap.clus %>% mutate(Cluster = as.factor(leiden.clusters)), aes(x = U1, y = U2, color = Cluster, shape = Cluster)) +
  geom_point() +
  theme_classic()
```

Expression profiles per cluster 

```{r fig.width=10, fig.height=10}
tcga.clustered <- tcga.raw %>%
  mutate(Cluster = as.factor(leiden.clusters)) %>%
  pivot_longer(names_to = "Marker", values_to = "Z", -Cluster)

profiles <- seq_len(max(leiden.clusters)) %>% map(~
ggplot(
  tcga.clustered %>% filter(Cluster == .x),
  aes(x = Marker, y = Z, color = Marker)
) +
  stat_summary(fun.data = mean_sdl, show.legend = FALSE) +
  ylim(-3, 3) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)))

plot_grid(plotlist = profiles, labels = paste("Cluster", seq_len(max(leiden.clusters))))
```

Differential expression analysis

```{r}
design <- model.matrix(~0 + as.factor(leiden.clusters))
colnames(design) <- paste0("Cluster", seq_len(max(leiden.clusters)))

tcga.leiden.limma <- lmFit(t(tcga.raw), design = design)

tcga.leiden.limma$coefficients

tests.leiden <- decideTests(eBayes(tcga.leiden.limma))

tests.leiden@.Data

summary(tests.leiden)
```


## Consensus NMF

```{r}

tcga.nmf.rank <- nmfEstimateRank(as.matrix(t(tcga)), seq(2,12), nrun = 10)

plot(tcga.nmf.rank)

tcga.nmf <- nmf(as.matrix(t(tcga)), 9, nrun = 10)

```

Extract basis of NMF (signature of cluster)

```{r}
basismap(tcga.nmf)
```

Extract coefficients of NMF (soft clustering of samples)

```{r}
coefmap(tcga.nmf)
```

Check for signs of overfitting

```{r}
consensusmap(tcga.nmf)
```


Assign clusters
```{r}
nmf.clusters <- apply(tcga.nmf@fit@H, 2, which.max)
```

Plot in 2D PCA and UMAP

```{r}
fviz_pca_ind(tcga.pca, geom = "point", col.ind = as.factor(nmf.clusters)) +
  theme_classic()

ggplot(tcga.umap.clus %>% mutate(Cluster = as.factor(nmf.clusters)), aes(x = U1, y = U2, color = Cluster)) +
  geom_point() +
  theme_classic()
```


Differential expression analysis

```{r}
design <- model.matrix(~0 + as.factor(nmf.clusters))
colnames(design) <- paste0("Cluster", seq_len(max(nmf.clusters)))

tcga.nmf.limma <- lmFit(t(tcga.raw), design = design)

tcga.nmf.limma$coefficients

tests.nmf <- decideTests(eBayes(tcga.nmf.limma))

tests.nmf@.Data

summary(tests.nmf)
```
