---
title: "Analysis of TCGA RPPA LIHC samples"
author: "Jovan Tanevski"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html:
    df_print: kable
editor_options:
  chunk_output_type: console
---

## Setup

Load required libraries.

```{r setup, message = FALSE}
library(tidyverse)
library(skimr)
library(uwot)
library(factoextra)
library(cowplot)
library(limma)
library(NMF)
library(pheatmap)
```

Read filtered TCGA RRPA data and display summary statistics.

```{r}
tcga.raw <- read_csv("data/TCGA-RPPA-LIHC_selected.csv", col_types = cols()) %>%
  select(-TumorType) %>%
  column_to_rownames("SampleID")

skim(tcga.raw)
```

Quantile normalization and rank normalization for NMF as suggested in https://gdac.broadinstitute.org/runs/analyses__2016_01_28/reports/cancer/LIHC/RPPA_Clustering_CNMF/nozzle.html

```{r}

tcga.norm <- normalizeQuantiles(tcga.raw)

skim(tcga.norm)

tcga.rank <- mutate_all(tcga.raw, rank)
```


## Dimensionality reduction

```{r}
tcga.pca <- prcomp(tcga.norm)
summary(tcga.pca)

set.seed(42)
tcga.umap <- umap(tcga.norm, n_neighbors = 10, n_epochs = 1000)
```


## Consensus NMF clustering

For this analysis 3 clusters were selected based on the consensus cophenetic correlation coefficient.

```{r}
tcga.nmf.rank <- nmfEstimateRank(as.matrix(t(tcga.rank)), seq(2,10), nrun = 20, seed = 42)

plot(tcga.nmf.rank)

tcga.nmf <- tcga.nmf <- tcga.nmf.rank$fit[["3"]]
```

Extract basis of NMF (signature of cluster)

```{r}
basismap(tcga.nmf)
```

Extract coefficients of NMF (soft clustering of samples)

```{r}
coefmap(tcga.nmf)
```

Check for signs of overfitting

```{r}
consensusmap(tcga.nmf)
```


Assign clusters
```{r}
nmf.clusters <- apply(tcga.nmf@fit@H, 2, which.max)
```

Plot in 2D PCA and UMAP

```{r}
fviz_pca_ind(tcga.pca, geom = "point", col.ind = as.factor(nmf.clusters)) +
  theme_classic()

tcga.umap.clus <-
  tcga.umap %>%
  cbind(nmf.clusters) %>%
  `colnames<-`(c("U1", "U2", "Cluster")) %>%
  as_tibble() %>%
  mutate_at("Cluster", as.factor)

ggplot(tcga.umap.clus, aes(x = U1, y = U2, color = Cluster)) +
  geom_point() +
  theme_classic()
```

Expression profiles per cluster 

```{r fig.width=10, fig.height=10}
tcga.clustered.nmf <- tcga.norm %>%
  mutate(Cluster = as.factor(nmf.clusters)) %>%
  pivot_longer(names_to = "Marker", values_to = "Norm.value", -Cluster)

profiles <- seq_len(max(nmf.clusters)) %>% map(~
ggplot(
  tcga.clustered.nmf %>% filter(Cluster == .x),
  aes(x = Marker, y = Norm.value, color = Marker)
) +
  stat_summary(fun.data = mean_sdl, show.legend = FALSE) +
  ylim(-3, 3) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)))

plot_grid(plotlist = profiles, labels = paste("Cluster", seq_len(max(nmf.clusters))))
```

Differential expression analysis (limma)

```{r}
design <- model.matrix(~0 + as.factor(nmf.clusters))
colnames(design) <- paste0("Cluster", seq_len(max(nmf.clusters)))

tcga.nmf.limma <- lmFit(t(tcga.norm), design = design)

tcga.nmf.eb <- eBayes(tcga.nmf.limma)

topTable(tcga.nmf.eb)

tests.nmf <- decideTests(tcga.nmf.eb)

tests.nmf@.Data

summary(tests.nmf)
```

Differential expression analysis (silhouette)

Calculate the similarity of samples using the expression and the silhouette scores based on the assigned clusters.
```{r}
silhouette.nmf <- silhouette(nmf.clusters, dist(tcga.norm))
fviz_silhouette(silhouette.nmf)
```

Select only the samples with positive silhouette scores as "core samples"
```{r}
core.samples <- which(silhouette.nmf[,3] > 0)
tcga.core.samples <- tcga.norm %>% add_column(Cluster = nmf.clusters) %>% 
  slice(core.samples)
```

Calculate difference in means (mean(cluster) - mean(other)), one-vs-all t-test per marker and correct for FDR. Filter q <= 0.05. Plot the differences.

```{r}
de.table <- unique(tcga.core.samples$Cluster) %>% map_dfr(\(c){
  tcga.core.samples %>% 
    summarize(across(-Cluster, ~t.test(.x ~ (Cluster==c))$p.value)) %>%
    pivot_longer(names_to = "Marker", values_to = "p", everything()) %>% 
    mutate(Cluster = c, Difference = tcga.core.samples %>% 
             group_by(Cluster == c) %>% 
             select(-Cluster) %>% 
             group_split(.keep = FALSE) %>% map(colMeans) %>% reduce(`-`))
 }) %>% 
  mutate(q = p.adjust(p, method = "fdr"), Difference = -Difference)

de.table %>%
  filter(q <= 0.05) %>%
  arrange(q)

de.table %>%
  pivot_wider(names_from = "Cluster", values_from = "Difference", -c(p, q)) %>%
  column_to_rownames("Marker") %>%
  as.matrix() %>%
  pheatmap(scale = "none")
```
